---
title: 'Diversity of *Scaptotrigona Mexicana* and *Melipona beecheii*'
subtitle: 'Reads'
author:
 - Adriana Haydé Contreras Peruyero (haydeeperuyero@gmail.com)
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r, echo=FALSE}
library("phyloseq")
library("ggplot2")
library("RColorBrewer")
library("patchwork")
library(readr)
library(tidyverse)
library(grid)
library(gridExtra)
```

First, we need to read the biom files. In this file, we found the following **Kingdoms**.

```{r, echo=FALSE}
raw_metagenomes <- import_biom("D:/Users/hayde/Documents/miel/miel/Metagenomics/data/bioms/db_levaduras/honey.biom")
raw_metagenomes@tax_table@.Data <- substring(raw_metagenomes@tax_table@.Data, 4)
colnames(raw_metagenomes@tax_table@.Data)<- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")


#raw_scapto <- import_biom("D:/Users/hayde/Documents/miel/miel/Metagenomics/data/bioms/db_levaduras/Scapto.biom")
#raw_scapto@tax_table@.Data <- substring(raw_scapto@tax_table@.Data, 4)
#colnames(raw_scapto@tax_table@.Data)<- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")


#raw_melli <- import_biom("D:/Users/hayde/Documents/miel/miel/Metagenomics/data/bioms/db_levaduras/Melli.biom")
#raw_melli@tax_table@.Data <- substring(raw_melli@tax_table@.Data, 4)
#colnames(raw_melli@tax_table@.Data)<- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")

```


```{r, include=FALSE, message=FALSE, warning=FALSE}
df_meta <- read_csv("D:/Users/hayde/Documents/miel/miel/Metagenomics/data/metadata_miel.csv")

#names(df_meta)[names(df_meta) == "Zootechnic Mangement"] <- "Zootechnic_Mangement"
#names(df_meta)[names(df_meta) == "Flowering type"] <- "Flowering_type"
#names(df_meta)[names(df_meta) == "Localidad"] <- "Locality"
```

```{r, include=FALSE}
df_meta <- df_meta %>%
  mutate(SampleID = case_when(
    grepl("_Scapto", ID) ~ paste0("Scapto_", sub("_Scapto.*", "", ID)),
    grepl("_Melli", ID) ~ paste0("Meli_", sub("_Melli.*", "", ID)),
    TRUE ~ as.character(ID)  # Para cualquier otro caso, mantener el valor original
  ))
```


```{r, eval=FALSE}
#df_meta_scapto <- df_meta %>%
#  filter(Specie == "Scaptotrigona Mexicana")

#df_meta_melli <- df_meta %>%
#  filter(Specie == "Melipona beecheii")
```


```{r}
df_metadata <- sample_data(df_meta)
rownames(df_metadata) <- df_meta$ID

#df_metadata_scapto <- sample_data(df_meta_scapto)
#rownames(df_metadata_scapto) <- df_meta_scapto$ID

#df_metadata_melli <- sample_data(df_meta_melli)
#rownames(df_metadata_melli) <- df_meta_melli$ID
```


```{r, echo=FALSE}
all_data <- merge_phyloseq(raw_metagenomes, sample_data(df_metadata))

#all_data_scapto <- merge_phyloseq(raw_scapto, sample_data(df_metadata_scapto))

#all_data_melli <- merge_phyloseq(raw_melli, sample_data(df_metadata_melli))
```

```{r, echo=FALSE}
# Obtener los nombres de las muestras
sample_names <- sample_names(all_data)


# Cambiar nombres con "_Melli" a "M_número"
sample_names <- gsub("(\\d+)_Melli", "Meli_\\1", sample_names)

# Cambiar nombres con "_Scapto" a "S_número"
sample_names <- gsub("(\\d+)_Scapto", "Scapto_\\1", sample_names)

# Asignar los nuevos nombres de las muestras al objeto phyloseq
sample_names(all_data) <- sample_names

# Verificar los nuevos nombres de las muestras
sample_names(all_data)


# Obtener los nombres de las muestras
#sample_names_scapto <- sample_names(all_data_scapto)

# Cambiar nombres con "_Scapto" a "S_número"
#new_sample_names_scapto <- gsub("(\\d+)_Scapto", "S_\\1", sample_names_scapto)

# Asignar los nuevos nombres de las muestras al objeto phyloseq
#sample_names(all_data_scapto) <- new_sample_names_scapto

# Verificar los nuevos nombres de las muestras
#sample_names(all_data_scapto)


# Obtener los nombres de las muestras
#sample_names_melli <- sample_names(all_data_melli)

# Cambiar nombres con "_Melli" a "M_número"
#new_sample_names_melli <- gsub("(\\d+)_Melli", "M_\\1", sample_names_melli)

# Asignar los nuevos nombres de las muestras al objeto phyloseq
#sample_names(all_data_melli) <- new_sample_names_melli

# Verificar los nuevos nombres de las muestras
#sample_names(all_data_melli)
```

```{r}
tax_table <- as.data.frame(tax_table(all_data))
reino_col <- "Kingdom"  # Ajusta el nombre de la columna si es diferente
reino_counts <- table(tax_table[[reino_col]])
total_observaciones <- sum(reino_counts)
porcentajes <- (reino_counts / total_observaciones) * 100
porcentajes

```


```{r, echo=FALSE}
dprof <- data.frame(Samples = colnames(all_data@otu_table@.Data),
                    Reads = sample_sums(all_data),
                    Specie = all_data@sam_data@.Data[[15]])

#dprof_scapto <- data.frame(Samples = colnames(all_data_scapto@otu_table@.Data),
#                    Reads = sample_sums(all_data_scapto),
#                   Bee = all_data_scapto@sam_data@.Data[[15]])

#dprof_melli <- data.frame(Samples = colnames(all_data_melli@otu_table@.Data),
#                    Reads = sample_sums(all_data_melli),
#                    Bee = all_data_melli@sam_data@.Data[[15]])

```



## Deep reads

In the following plot, we can see the deep of the reads for each sample in the honey of *Scaptotrigona mexicana* specie.

```{r, fig.dim=(12,8)   echo=FALSE}
reads <- ggplot(data = dprof, mapping = aes(x = Samples, y = Reads))+
  geom_bar(stat = "identity", aes( fill = Specie)) +
  scale_fill_manual(values=c("#b8860b", "#FFBF00"))+
  facet_wrap(~Specie, scales = "free") +
  labs(title = "Deep")+
  theme(axis.text.x = element_text(angle = 55, hjust = 1))+
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "bold"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14, face = "italic"), # Tamaño del texto de la leyenda
    legend.position = "bottom"
  )
  
reads
```


```{r, include=FALSE}
#ggsave(filename = "~/miel/Metagenomics/figures/deep_Scapto.svg", plot = reads, width = 8, height = 5, dpi = 300)

ggsave(filename = "D:/Users/hayde/Documents/miel/miel/Metagenomics/figures/figs/deep_reads3.svg", plot = reads, width = 10, height = 8, dpi = 300)
```

First, we will work with the `Bacteria` subset and then with the `Viruses`.


# Bacteria

A summary of the Bacteria subset:


```{r, echo=FALSE}
# bacterias
dataB <- subset_taxa(all_data, Kingdom == "Bacteria")
sample_sums(dataB)
summary(dataB@otu_table@.Data)

#dataB_S <- subset_taxa(all_data_scapto, Kingdom == "Bacteria")
#sample_sums(dataB_S)
#summary(dataB_S@otu_table@.Data)

#dataB_M <- subset_taxa(all_data_melli, Kingdom == "Bacteria")
#sample_sums(dataB_M)
#summary(dataB_M@otu_table@.Data)
```


## Alpha diversity 

- **Observed:** The Observed index simply counts the number of unique species (or Operational Taxonomic Units, OTUs) in a sample. It is a basic measure of species richness.

**Formula:**  
\[ S_{\text{obs}} = \sum_{i=1}^S I(n_i > 0) \]  
where \( S \) is the total number of species and \( I(n_i > 0) \) is an indicator function that is 1 if species \( i \) is observed (i.e., \( n_i > 0 \)) and 0 otherwise.

- **Chao1:** The Chao1 index is an estimator of the true species richness in a sample, accounting for the number of rare species. It is particularly useful when dealing with undersampled communities.

**Formula:**  
\[ \text{Chao1} = S_{\text{obs}} + \frac{F_1(F_1 - 1)}{2(F_2 + 1)} \]  
where \( F_1 \) is the number of singletons (species observed exactly once) and \( F_2 \) is the number of doubletons (species observed exactly twice).

- **Shannon:** The Shannon index, also known as Shannon-Wiener or Shannon-Weaver index, measures the diversity in a community by considering both the richness and the evenness of species. It accounts for the abundance of each species.

**Formula:**  
\[ H' = -\sum_{i=1}^S p_i \ln(p_i) \]  
where \( S \) is the total number of species and \( p_i \) is the proportion of individuals belonging to species \( i \).

In the case of the samples of honey, we can observed the following with respect to the diversity of these indexes. 

The Observed index counts the number of unique species in each sample. The plot shows that:
- Sample `23S` has the lowest observed species richness.
- Sample `92S` has the highest observed species richness.
- There is variability in species richness among the samples, with some samples having significantly higher species richness compared to others.

The Chao1 index estimates the true species richness, accounting for rare species. The plot indicates that:
- Sample `23S` has the lowest estimated species richness.
- Sample `92S` has the highest estimated species richness, similar to the Observed index.
- The error bars represent the uncertainty in the Chao1 estimates, showing that some samples have a higher level of uncertainty.
- The pattern of richness among samples is consistent with the Observed index, but Chao1 generally provides a higher estimate due to its correction for unseen species.

The Shannon index measures both species richness and evenness. The plot shows that:
- Sample `93S` has the lowest Shannon diversity, indicating either lower richness, lower evenness, or both.
- Sample `23S` has the highest Shannon diversity, suggesting higher richness, evenness, or both.
- The variability in Shannon diversity across samples is smaller compared to Observed and Chao1 indices, suggesting that evenness plays a significant role in these diversity measures.

In general, we can see the following:

- The samples exhibit different levels of species richness and diversity.
- Samples `23S` and `92S` stand out, with `23S` having the lowest richness but highest evenness, and `92S` having the highest richness but moderate evenness.
- Chao1 provides a more conservative estimate of species richness by accounting for rare species, which is reflected in the wider range and error bars.
- The Shannon index suggests that species evenness varies across samples, impacting the overall diversity.


```{r, echo=FALSE}
alpha_plot_observed <- plot_richness(physeq = dataB, color= "Locality",
                                title = "Alpha diversity",
                                measures = c("Observed"))+
  facet_grid(~Specie, scales = "free")+
  scale_color_brewer(palette = "Dark2") +
  theme(strip.text = element_text(face = "italic"))+
  labs(x= "Samples")+
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "bold"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14), # Tamaño del texto de la leyenda
    legend.position = "bottom"
    )
alpha_plot_observed
```
```{r, echo=FALSE}
alpha_plot_shanon <- plot_richness(physeq = dataB, color= "Locality",
                                title = "Alpha diversity",
                                measures = c("Shannon"))+
  facet_grid(~Specie, scales = "free")+
  scale_color_brewer(palette = "Dark2") +
  theme(strip.text = element_text(face = "italic"))+
  labs(x= "Samples")+
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "bold"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14), # Tamaño del texto de la leyenda
    legend.position = "bottom"
    )
alpha_plot_shanon
```

```{r, echo=FALSE}
alpha_plot_sup <- plot_richness(physeq = dataB, color= "Locality",
                                title = "Alpha diversity",
                                measures = c("Observed", "Shannon"))+
  #facet_grid(~Specie, scales = "free")+
  #theme(strip.text = element_text(face = "italic"))+
  scale_color_brewer(palette = "Dark2") +
  labs(x= "Samples")+
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "bold"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14), # Tamaño del texto de la leyenda
    legend.position = "bottom"
  )
alpha_plot_sup
```

```{r, eval=FALSE,echo=FALSE}

## Ya no
alpha_scapto <- plot_richness(physeq = dataB_S, 
                                title = "Scaptotrigona mexicana",
                                measures = c("Observed","Chao1","Shannon"))#+
  #scale_color_manual(values = c("blue", "darkblue"))

alpha_melli <- plot_richness(physeq = dataB_M, 
                                title = "Melipona beecheii",
                                measures = c("Observed","Chao1","Shannon"))
  
#alpha_div_plot <- plot_richness(physeq = dataB, 
#                                title = "Alpha diversity indexes",
#                                measures = c("Chao1"))

# Personalizar el gráfico
alpha_scapto <- alpha_scapto + 
  labs(
    x = "Samples"
  ) +
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "italic"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14) # Tamaño del texto de la leyenda
  )
  #scale_color_manual(values = c("Observed" = "blue", "Chao1" = "darkblue", "Shannon" = "lightblue")) +
  #scale_fill_manual(values = c("Observed" = "blue", "Chao1" = "darkblue", "Shannon" = "lightblue"))



alpha_melli <- alpha_melli + 
  labs(
    x = "Samples"
  ) +
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "italic"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_blank(), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14) # Tamaño del texto de la leyenda
  )


title <- textGrob("Alpha diversity indexes", gp = gpar(fontsize = 20, fontface = "bold"))

# Colocar los gráficos juntos con un título general usando grid.arrange
alpha_plot <- grid.arrange(
  arrangeGrob(alpha_scapto, alpha_melli, ncol = 2),  # Colocar los gráficos en una fila
  top = title  # Agregar el título general
)


#grid.arrange( ncol=2)
```

```{r, include=FALSE}
#ggsave(filename = "~/miel/Metagenomics/figures/alpha_div_plot_Scapto.svg", plot = alpha_plot, width = 8, height = 5, dpi = 300)

#ggsave(filename = "D:/Users/hayde/Documents/miel/miel/Metagenomics/figures/figs/alpha_div_plot.png", plot = alpha_plot, width = 7, height = 5, dpi = 300)

ggsave(filename = "D:/Users/hayde/Documents/miel/miel/Metagenomics/figures/figs/alpha_div_plot_sup2.svg", plot = alpha_plot_sup, width = 7, height = 5, dpi = 300)
```

```{r, include=FALSE, echo=FALSE}
summary(dataB@tax_table@.Data== "")
raw_metagenomesB <- subset_taxa(dataB, Genus != "")
summary(dataB@tax_table@.Data== "")

# summary(dataB_S@tax_table@.Data== "")
# raw_metagenomesB_S <- subset_taxa(dataB_S, Genus != "")
# summary(dataB_S@tax_table@.Data== "")
# 
# summary(dataB_M@tax_table@.Data== "")
# raw_metagenomesB_M <- subset_taxa(dataB_M, Genus != "")
# summary(dataB_M@tax_table@.Data== "")
```


## Beta diversity


Beta diversity measures the differences in species composition between samples. The following plot is an NMDS (Non-metric Multidimensional Scaling) plot, which is commonly used to visualize beta diversity. 

### Interpretation of the NMDS Plot

1. **Axes Interpretation**:
   - **NMDS1** and **NMDS2** represent the two dimensions that best capture the differences in species composition among the samples.
   - The scales on these axes are arbitrary and do not have direct biological meaning. Instead, the relative positions of the points are what matter.

2. **Points Representation**:
   - Each point on the plot represents a sample.
   - The closer two points are to each other, the more similar their species compositions are.
   - Conversely, points that are farther apart indicate samples with more dissimilar species compositions.

3. **Clustering**:
   - If points form clusters, it indicates groups of samples that have similar species compositions.
   - In this plot, there seems to be some clustering, which suggests that certain samples have more similar communities.

4. **Stress Value (not shown)**:
   - Normally, NMDS plots include a stress value, which indicates how well the two-dimensional plot represents the multi-dimensional data. Lower stress values (typically <0.2) suggest a good representation.
   - Since the stress value is not shown here, it's important to be cautious in interpreting the distances, but generally, NMDS plots provide a useful visual summary.

### Overall Interpretation

- The NMDS plot shows the beta diversity among the samples, highlighting similarities and differences in their species compositions.
- Samples that are close to each other on the plot have similar species compositions, while those that are far apart are more dissimilar.
- Identifying clusters or gradients in the plot can help infer patterns related to environmental gradients, treatments, or other factors influencing the community composition.


In the plot we can see that there are some correlation between the samples and the city of the samples.

```{r, include=FALSE}
#Bacteria
rel_dataB <- transform_sample_counts(dataB, function(x) x*100 / sum(x) )
```

```{r, echo=FALSE}
meta_ord <- ordinate(physeq = rel_dataB, method = "NMDS", 
                     distance = "bray")

beta_div_plot <- plot_ordination(physeq = rel_dataB, ordination = meta_ord, color="Locality", shape="Specie") +
  geom_point(size = 4) +  # Adjust point size
  #geom_text(aes(label = SampleID), vjust = -1.5) +  # Add labels above points
  scale_color_brewer(palette = "Dark2") +  # Paleta predefinida
  theme_minimal() +  # Use a minimal theme for clean look
  theme(legend.position = "right") +  # Position legend on the right
  labs(title = "Beta Diversity NMDS", x = "NMDS1", y = "NMDS2")+
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "bold"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14, face = "italic") # Tamaño del texto de la leyenda
  )

beta_div_plot

```



```{r, include=FALSE}
#ggsave(filename = "~/miel/Metagenomics/figures/beta_div_plot_Scapto.svg", plot = beta_div_plot, width = 8, height = 8, dpi = 300)

ggsave(filename = "D:/Users/hayde/Documents/miel/miel/Metagenomics/figures/figs/beta_div_plot.svg", plot = beta_div_plot, width = 8, height = 6, dpi = 300)
```


## Absolute abundance

```{r, include=FALSE}
absolute_phylum_global <- tax_glom(physeq = dataB, taxrank = "Phylum")
absolute_phyl_global_df <- psmelt(absolute_phylum_global)
str(absolute_phyl_global_df)

absolute_phyl_global_df$Phylum <- as.factor(absolute_phyl_global_df$Phylum)
phylum_colors_abs<- colorRampPalette(brewer.pal(8,"Dark2")) (length(levels(absolute_phyl_global_df$Phylum)))
```

```{r, echo=FALSE}
absolute_plot <- ggplot(data= absolute_phyl_global_df, aes(x=Sample, y=Abundance, fill=Phylum))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = phylum_colors_abs)+
  labs(x= "Samples", y = "Absolute abundance") +
  ggtitle("Absolute abundance")+
  theme(legend.text = element_text(size = 6), legend.key.size = unit(0.3, 'cm'), axis.title = element_text(size = 10), axis.text.x = element_text(size = 8, angle = 90), plot.title = element_text(size = 14))


absolute_plot
```

## Relative abundance


```{r, include=FALSE}
rel_phylum_global <- tax_glom(rel_dataB, taxrank = 'Phylum')
rel_phyl_global_df<-psmelt(rel_phylum_global) # convert to data frame
str(rel_phyl_global_df)
rel_phyl_global_df$Phylum <- as.factor(rel_phyl_global_df$Phylum)
phylum_colors_rel<- colorRampPalette(brewer.pal(8,"Dark2")) (length(levels(rel_phyl_global_df$Phylum)))
```


```{r, echo=FALSE}
relative_plot <- ggplot(data=rel_phyl_global_df, aes(x=Sample, y=Abundance, fill=Phylum))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = phylum_colors_rel)+
  labs(x= "Samples", y = "Relative abundance") +
  ggtitle("Relative abundance")+
  theme(legend.text = element_text(size = 6), legend.key.size = unit(0.3, 'cm'), axis.title = element_text(size = 10), axis.text.x = element_text(size = 10, angle = 90), plot.title = element_text(size = 14))

relative_plot

```




```{r, include=FALSE}
#Subset
rel_phylum_global <- tax_glom(dataB, taxrank = 'Phylum')
rel_phylum_global <- transform_sample_counts(rel_phylum_global, function(x) x*100 / sum(x) )
rel_phyl_global_df<-psmelt(rel_phylum_global) # convert to data frame
str(rel_phyl_global_df)
rel_phyl_global_df$Phylum[rel_phyl_global_df$Abundance < 5] <- "Phyla < 5%abund"
rel_phyl_global_df$Phylum <- as.factor(rel_phyl_global_df$Phylum)
#phylum_colors_rel<- colorRampPalette(brewer.pal(8,"Dark2")) (length(levels(rel_phyl_global_df$Phylum)))
```

```{r, echo=FALSE}
# Define your custom color palette
phylum_colors_rel <- c(
  "Bacillota" = "#1f78b4", 
  "Pseudomonadota" = "#33a02c",
  "Phyla < 5%abund" = "#6a3d9a"
  # Add other phyla and their corresponding colors
)


relative_plot_f <- ggplot(data=rel_phyl_global_df, aes(x=Sample, y=Abundance, fill=Phylum))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = phylum_colors_rel)+
  
  #facet_grid(~City)+
  labs(x= "Samples", y = "Relative abundance") +
  ggtitle("Relative abundance")+
  theme(legend.text = element_text(size = 16), legend.key.size = unit(0.4, 'cm'), axis.title = element_text(size = 16), axis.text.x = element_text(size = 16, angle = 90),axis.text.y = element_text(size = 16), plot.title = element_text(size = 20), axis.title.x = element_text(size = 16, face = "bold"), 
    axis.title.y = element_text(size = 16, face = "bold"), legend.title = element_text(size = 16),
    legend.position = "bottom")

relative_plot_f
```

```{r}
Melipona_bacillota <- rel_phyl_global_df %>%
  filter(Specie == "Melipona beecheii", Phylum == "Bacillota") %>%  # Filtrar las muestras de abeja y el phylum
  group_by(Sample) %>%  # Agrupar por muestra
  mutate(total_abundance = sum(Abundance))
# Mostrar el resultado
summary(Melipona_bacillota$Abundance)

Scaptotrigona_Bacillota <- rel_phyl_global_df %>%
  filter(Specie == "Scaptotrigona mexicana", Phylum == "Bacillota") %>%  # Filtrar las muestras de abeja y el phylum
  group_by(Sample) %>%  # Agrupar por muestra
  mutate(total_abundance = sum(Abundance))
# Mostrar el resultado
summary(Scaptotrigona_Bacillota$Abundance)


Melipona_Pseudomonadota <- rel_phyl_global_df %>%
  filter(Specie == "Melipona beecheii", Phylum == "Pseudomonadota") %>%  # Filtrar las muestras de abeja y el phylum
  group_by(Sample) %>%  # Agrupar por muestra
  mutate(total_abundance = sum(Abundance))
# Mostrar el resultado
summary(Melipona_Pseudomonadota$Abundance)

Scaptotrigona_Pseudomonadota <- rel_phyl_global_df %>%
  filter(Specie == "Scaptotrigona mexicana", Phylum == "Pseudomonadota") %>%  # Filtrar las muestras de abeja y el phylum
  group_by(Sample) %>%  # Agrupar por muestra
  mutate(total_abundance = sum(Abundance))
# Mostrar el resultado
summary(Scaptotrigona_Pseudomonadota$Abundance)

```


```{r, include=FALSE}
ggsave(filename = "D:/Users/hayde/Documents/miel/miel/Metagenomics/figures/figs/relative_scapto_phylum.svg", plot = relative_plot_f, width = 7, height = 5, dpi = 300)
#ggsave(filename = "~/miel/Metagenomics/figures/relative_scapto_phylum.svg", plot = relative_plot_f, width = 6, height = 5, dpi = 300)
```


```{r, include=FALSE, eval=FALSE}
rel_Genus_global <- tax_glom(rel_dataB, taxrank = 'Genus')
rel_Genus_global_df<-psmelt(rel_Genus_global) # convert to data frame
str(rel_Genus_global_df)
rel_Genus_global_df$Genus <- as.factor(rel_Genus_global_df$Genus)
Genus_colors_rel<- colorRampPalette(brewer.pal(8,"Dark2")) (length(levels(rel_Genus_global_df$Genus)))
```


```{r, include=FALSE, echo=FALSE, eval=FALSE}
relative_plot <- ggplot(data=rel_Genus_global_df, aes(x=Sample, y=Abundance, fill=Genus))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = Genus_colors_rel)+
  labs(x= "Samples", y = "Relative abundance") +
  ggtitle("Relative abundance")+
  theme(legend.text = element_text(size = 6), legend.key.size = unit(0.3, 'cm'), axis.title = element_text(size = 10), axis.text.x = element_text(size = 10, angle = 90), plot.title = element_text(size = 14), legend.position = "bottom")

relative_plot

```




```{r, include=FALSE}
#Subset
rel_Genus_global <- tax_glom(dataB, taxrank = 'Genus')
rel_Genus_global <- transform_sample_counts(rel_Genus_global, function(x) x*100 / sum(x) )
rel_Genus_global_df<-psmelt(rel_Genus_global) # convert to data frame
str(rel_Genus_global_df)
rel_Genus_global_df$Genus[rel_Genus_global_df$Abundance < 5] <- "Genus < 5% abund"
rel_Genus_global_df$Genus <- as.factor(rel_Genus_global_df$Genus)

```

```{r, echo=FALSE}
#Genus_colors_rel<- colorRampPalette(brewer.pal(10,"Dark2")) (length(levels(rel_Genus_global_df$Genus)))

#genus_colors <- c(
#  "Acetilactobacillus" = "#1f78b4", 
#  "Apilactobacillus" = "#33a02c",
#  "Lactobacillus" = "#e31a1c", 
#  "Nicoliella" = "#ff7f00",
#  "Pantoea" = "#ff007f",
#  "Tatumella" = "#b15928",
#  "Fructobacillus" = "#ffbf00",
#  "Genus < 5% abund" = "#6a3d9a"
#)


genus_colors <- c(
  "Acetilactobacillus" = "#002868", 
  "Apilactobacillus" = "#0077C2",
  "Lactobacillus" = "#66B2FF", 
  "Nicoliella" = "#ADD8E1",
  "Pantoea" = "#008080",
  "Tatumella" = "#98FB98",
  "Fructobacillus" = "#007BA7",
  "Genus < 5% abund" = "#778899"
)

rel_Genus_global_df$Genus <- factor(rel_Genus_global_df$Genus, levels = c("Acetilactobacillus", "Apilactobacillus", "Fructobacillus", "Lactobacillus", "Nicoliella", "Pantoea", "Tatumella","Genus < 5% abund" ))

relative_plot_g <- ggplot(data=rel_Genus_global_df, aes(x=Sample, y=Abundance, fill=Genus))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = genus_colors)+
  #facet_wrap(~Locality, drop = TRUE, scales = "free")+
  labs(x= "Samples", y = "Relative abundance") +
  ggtitle("Relative abundance")+
   theme(#strip.text = element_text(size = 16, face = "bold"),
         legend.text = element_text(size = 16), 
         legend.key.size = unit(0.4, 'cm'), 
         axis.title = element_text(size = 16), 
         axis.text.x = element_text(size = 16, angle = 90),
         axis.text.y = element_text(size = 16), 
         plot.title = element_text(size = 20), 
         axis.title.x = element_text(size = 16, face = "bold"),
         axis.title.y = element_text(size = 16, face = "bold"), 
         legend.title = element_text(size = 16),
         legend.position = "bottom")

relative_plot_g
```

```{r, include=FALSE}
ggsave(filename = "D:/Users/hayde/Documents/miel/miel/Metagenomics/figures/figs/relative_scapto_genus_25.svg", plot = relative_plot_g, width = 10, height = 8, dpi = 300)
#ggsave(filename = "~/miel/Metagenomics/figures/relative_scapto_genus.svg", plot = relative_plot_g, width = 8, height = 7, dpi = 300)

```



# Virus

```{r, echo=FALSE}
# virus
dataV <- subset_taxa(all_data, Kingdom == "Viruses")
sample_sums(dataV)
summary(dataV@otu_table@.Data)

```



## Alfa diversidad


```{r, echo=FALSE}
alpha_div_plot <- plot_richness(physeq = dataV, 
                                title = "Alpha diversity indexes Viruses",
                                measures = c("Observed","Chao1","Shannon")) 

alpha_div_plot
```

```{r, echo=FALSE}
summary(dataV@tax_table@.Data== "")
raw_metagenomesV <- subset_taxa(dataV, Genus != "")
summary(dataV@tax_table@.Data== "")
```


## Diversidad

```{r, echo=FALSE}
#Virus
#Abundancias relativas
rel_dataV <- transform_sample_counts(dataV, function(x) x*100 / sum(x) )
```


### Diversidad Beta

```{r, echo=FALSE}
meta_ord <- ordinate(physeq = rel_dataV, method = "NMDS", 
                     distance = "bray")

beta_div_plot <- plot_ordination(physeq = rel_dataV, ordination = meta_ord)

beta_div_plot

```


### Diversidad absoluta

```{r, echo=FALSE}
absolute_phylum_global <- tax_glom(physeq = dataV, taxrank = "Phylum")
absolute_phyl_global_df <- psmelt(absolute_phylum_global)
str(absolute_phyl_global_df)

absolute_phyl_global_df$Phylum <- as.factor(absolute_phyl_global_df$Phylum)
phylum_colors_abs<- colorRampPalette(brewer.pal(8,"Dark2")) (length(levels(absolute_phyl_global_df$Phylum)))
```

```{r, echo=FALSE}
absolute_plot <- ggplot(data= absolute_phyl_global_df, aes(x=Sample, y=Abundance, fill=Phylum))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = phylum_colors_abs)+
  labs(x= "Samples", y = "Absolute abundance") +
  ggtitle("Absolute abundance")+
  theme(legend.text = element_text(size = 6), legend.key.size = unit(0.3, 'cm'), axis.title = element_text(size = 10), axis.text.x = element_text(size = 8, angle = 90), plot.title = element_text(size = 14))

absolute_plot
```

### Diversidad relativa


```{r, echo=FALSE}
rel_phylum_global <- tax_glom(rel_dataV, taxrank = 'Phylum')
rel_phyl_global_df<-psmelt(rel_phylum_global) # convert to data frame
str(rel_phyl_global_df)
rel_phyl_global_df$Phylum <- as.factor(rel_phyl_global_df$Phylum)
phylum_colors_rel<- colorRampPalette(brewer.pal(8,"Dark2")) (length(levels(rel_phyl_global_df$Phylum)))
```


```{r, echo=FALSE}
relative_plot <- ggplot(data=rel_phyl_global_df, aes(x=Sample, y=Abundance, fill=Phylum))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = phylum_colors_rel)+
  labs(x= "Samples", y = "Relative abundance") +
  ggtitle("Relative abundance")+
  theme(legend.text = element_text(size = 6), legend.key.size = unit(0.3, 'cm'), axis.title = element_text(size = 10), axis.text.x = element_text(size = 10, angle = 90), plot.title = element_text(size = 14))

relative_plot

```




```{r, echo=FALSE}
#Subset
rel_phylum_global <- tax_glom(rel_dataV, taxrank = 'Phylum')
rel_phyl_global_df<-psmelt(rel_phylum_global) # convert to data frame
str(rel_phyl_global_df)
rel_phyl_global_df$Phylum[rel_phyl_global_df$Abundance < 0.5] <- "Phyla < 0.5% abund."
rel_phyl_global_df$Phylum <- as.factor(rel_phyl_global_df$Phylum)
phylum_colors_rel<- colorRampPalette(brewer.pal(8,"Dark2")) (length(levels(rel_phyl_global_df$Phylum)))
```

```{r, echo=FALSE}
relative_plot_f <- ggplot(data=rel_phyl_global_df, aes(x=Sample, y=Abundance, fill=Phylum))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = phylum_colors_rel)+
  labs(x= "Samples", y = "Relative abundance") +
  ggtitle("Relative abundance")+
  theme(legend.text = element_text(size = 6), legend.key.size = unit(0.3, 'cm'), axis.title = element_text(size = 10), axis.text.x = element_text(size = 10, angle = 90), plot.title = element_text(size = 14))

relative_plot_f

```




```{r, echo=FALSE}
absolute_Genus_global <- tax_glom(physeq = dataV, taxrank = "Genus")
absolute_Genus_global_df <- psmelt(absolute_Genus_global)
str(absolute_Genus_global_df)

absolute_Genus_global_df$Genus <- as.factor(absolute_Genus_global_df$Genus)
Genus_colors_abs<- colorRampPalette(brewer.pal(8,"Dark2")) (length(levels(absolute_Genus_global_df$Genus)))
```

```{r, echo=FALSE}
absolute_plot <- ggplot(data= absolute_Genus_global_df, aes(x=Sample, y=Abundance, fill=Genus))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = Genus_colors_abs)+
  labs(x= "Samples", y = "Absolute abundance") +
  ggtitle("Absolute abundance")+
  theme(legend.text = element_text(size = 6), legend.key.size = unit(0.3, 'cm'), axis.title = element_text(size = 10), axis.text.x = element_text(size = 8, angle = 90), plot.title = element_text(size = 14))

absolute_plot
```




```{r, echo=FALSE}
rel_Genus_global <- tax_glom(rel_dataV, taxrank = 'Genus')
rel_Genus_global_df<-psmelt(rel_Genus_global) # convert to data frame
str(rel_Genus_global_df)
rel_Genus_global_df$Genus <- as.factor(rel_Genus_global_df$Genus)
Genus_colors_rel<- colorRampPalette(brewer.pal(8,"Dark2")) (length(levels(rel_Genus_global_df$Genus)))
```


```{r, echo=FALSE}
relative_plot <- ggplot(data=rel_Genus_global_df, aes(x=Sample, y=Abundance, fill=Genus))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = Genus_colors_rel)+
  labs(x= "Samples", y = "Relative abundance") +
  ggtitle("Relative abundance")+
  theme(legend.text = element_text(size = 6), legend.key.size = unit(0.3, 'cm'), axis.title = element_text(size = 10), axis.text.x = element_text(size = 10, angle = 90), plot.title = element_text(size = 14))

relative_plot

```



```{r, echo=FALSE}
#Subset
rel_Genus_global <- tax_glom(dataV, taxrank = 'Genus')
rel_Genus_global <- transform_sample_counts(rel_Genus_global, function(x) x*100 / sum(x) )
rel_Genus_global_df<-psmelt(rel_Genus_global) # convert to data frame
str(rel_Genus_global_df)
rel_Genus_global_df$Genus[rel_Genus_global_df$Abundance < 5] <- "Genus < 5% abund"
rel_Genus_global_df$Genus <- as.factor(rel_Genus_global_df$Genus)
Genus_colors_rel<- colorRampPalette(brewer.pal(8,"Dark2")) (length(levels(rel_Genus_global_df$Genus)))
```

```{r, echo=FALSE}
relative_plot_g <- ggplot(data=rel_Genus_global_df, aes(x=Sample, y=Abundance, fill=Genus))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = Genus_colors_rel)+
  #scale_fill_manual(values = genus_colors)+
  #facet_wrap(~Locality, drop = TRUE, scales = "free")+
  labs(x= "Samples", y = "Relative abundance") +
  ggtitle("Relative abundance")+
   theme(#strip.text = element_text(size = 16, face = "bold"),
         legend.text = element_text(size = 16), 
         legend.key.size = unit(0.4, 'cm'), 
         axis.title = element_text(size = 16), 
         axis.text.x = element_text(size = 16, angle = 90),
         axis.text.y = element_text(size = 16), 
         plot.title = element_text(size = 20), 
         axis.title.x = element_text(size = 16, face = "bold"),
         axis.title.y = element_text(size = 16, face = "bold"), 
         legend.title = element_text(size = 16),
         legend.position = "bottom")

relative_plot_g


```

```{r, include=FALSE}
ggsave(filename = "D:/Users/hayde/Documents/miel/miel/Metagenomics/figures/figs/relative_scapto_genus_virus.svg", plot = relative_plot_g, width = 10, height = 8, dpi = 300)
#ggsave(filename = "~/miel/Metagenomics/figures/relative_scapto_genus.svg", plot = relative_plot_g, width = 8, height = 7, dpi = 300)

```



## Eu

```{r}
unique(all_data@tax_table@.Data[,"Kingdom"])
```

```{r}
dataE <- subset_taxa(all_data, Kingdom == "Eukaryota")
sample_sums(dataE)
summary(dataE@otu_table@.Data)
```

```{r}
dataFungi <- subset_taxa(dataE, Phylum == c("Ascomycota", "Basidiomycota", "Microsporidia"))
sample_sums(dataFungi)
summary(dataFungi@otu_table@.Data)
```


```{r}
#unique(dataE@tax_table@.Data[,"Phylum"])
#Ascomycota, Basidiomycota, Microsporidia

rel_phylum_global <- tax_glom(dataFungi, taxrank = 'Phylum')
rel_phylum_global <- transform_sample_counts(rel_phylum_global, function(x) x*100 / sum(x) )
rel_phyl_global_df<-psmelt(rel_phylum_global) # convert to data frame
str(rel_phyl_global_df)
#rel_phyl_global_df$Phylum[rel_phyl_global_df$Abundance < 5] <- "Phyla < 5%abund"
rel_phyl_global_df$Phylum <- as.factor(rel_phyl_global_df$Phylum)
phylum_colors_rel<- colorRampPalette(brewer.pal(8,"Dark2")) (length(levels(rel_phyl_global_df$Phylum)))
```

```{r, echo=FALSE}
# Define your custom color palette
#phylum_colors_rel <- c(
#  "Bacillota" = "#1f78b4", 
#  "Pseudomonadota" = "#33a02c",
#  "Phyla < 5%abund" = "#6a3d9a"
  # Add other phyla and their corresponding colors
#)


relative_plot_f <- ggplot(data=rel_phyl_global_df, aes(x=Sample, y=Abundance, fill=Phylum))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = phylum_colors_rel)+
  #facet_grid(~City)+
  labs(x= "Samples", y = "Relative abundance") +
  ggtitle("Relative abundance")+
  theme(legend.text = element_text(size = 16), legend.key.size = unit(0.4, 'cm'), axis.title = element_text(size = 16), axis.text.x = element_text(size = 16, angle = 90),axis.text.y = element_text(size = 16), plot.title = element_text(size = 20), axis.title.x = element_text(size = 16, face = "bold"), 
    axis.title.y = element_text(size = 16, face = "bold"), legend.title = element_text(size = 16),
    legend.position = "bottom")

relative_plot_f
```


```{r}
unique(dataE@tax_table@.Data[,"Phylum"])
#Ascomycota, Basidiomycota, Microsporidia

rel_phylum_global <- tax_glom(dataE, taxrank = 'Phylum')
rel_phylum_global <- transform_sample_counts(rel_phylum_global, function(x) x*100 / sum(x) )
rel_phyl_global_df<-psmelt(rel_phylum_global) # convert to data frame
str(rel_phyl_global_df)
#rel_phyl_global_df$Phylum[rel_phyl_global_df$Abundance < 5] <- "Phyla < 5%abund"
rel_phyl_global_df$Phylum <- as.factor(rel_phyl_global_df$Phylum)
phylum_colors_rel<- colorRampPalette(brewer.pal(8,"Dark2")) (length(levels(rel_phyl_global_df$Phylum)))
```

```{r, echo=FALSE}
# Define your custom color palette
#phylum_colors_rel <- c(
#  "Bacillota" = "#1f78b4", 
#  "Pseudomonadota" = "#33a02c",
#  "Phyla < 5%abund" = "#6a3d9a"
  # Add other phyla and their corresponding colors
#)


relative_plot_f <- ggplot(data=rel_phyl_global_df, aes(x=Sample, y=Abundance, fill=Phylum))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = phylum_colors_rel)+
  #facet_grid(~City)+
  labs(x= "Samples", y = "Relative abundance") +
  ggtitle("Relative abundance")+
  theme(legend.text = element_text(size = 16), legend.key.size = unit(0.4, 'cm'), axis.title = element_text(size = 16), axis.text.x = element_text(size = 16, angle = 90),axis.text.y = element_text(size = 16), plot.title = element_text(size = 20), axis.title.x = element_text(size = 16, face = "bold"), 
    axis.title.y = element_text(size = 16, face = "bold"), legend.title = element_text(size = 16),
    legend.position = "bottom")

relative_plot_f
```



```{r, echo=FALSE}
#Subset
rel_Genus_global <- tax_glom(dataE, taxrank = 'Genus')
rel_Genus_global <- transform_sample_counts(rel_Genus_global, function(x) x*100 / sum(x) )
rel_Genus_global_df<-psmelt(rel_Genus_global) # convert to data frame
str(rel_Genus_global_df)
rel_Genus_global_df$Genus[rel_Genus_global_df$Abundance < 1] <- "Genus < 1% abund"
rel_Genus_global_df$Genus <- as.factor(rel_Genus_global_df$Genus)
Genus_colors_rel<- colorRampPalette(brewer.pal(8,"Dark2")) (length(levels(rel_Genus_global_df$Genus)))
```

```{r, echo=FALSE}
relative_plot_g <- ggplot(data=rel_Genus_global_df, aes(x=Sample, y=Abundance, fill=Genus))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = Genus_colors_rel)+
  #scale_fill_manual(values = genus_colors)+
  #facet_wrap(~Locality, drop = TRUE, scales = "free")+
  labs(x= "Samples", y = "Relative abundance") +
  ggtitle("Relative abundance")+
   theme(#strip.text = element_text(size = 16, face = "bold"),
         legend.text = element_text(size = 16), 
         legend.key.size = unit(0.4, 'cm'), 
         axis.title = element_text(size = 16), 
         axis.text.x = element_text(size = 16, angle = 90),
         axis.text.y = element_text(size = 16), 
         plot.title = element_text(size = 20), 
         axis.title.x = element_text(size = 16, face = "bold"),
         axis.title.y = element_text(size = 16, face = "bold"), 
         legend.title = element_text(size = 16),
         legend.position = "bottom")

relative_plot_g


```

```{r, include=FALSE}
ggsave(filename = "D:/Users/hayde/Documents/miel/miel/Metagenomics/figures/figs/relative_scapto_genus_virus.svg", plot = relative_plot_g, width = 10, height = 8, dpi = 300)
#ggsave(filename = "~/miel/Metagenomics/figures/relative_scapto_genus.svg", plot = relative_plot_g, width = 8, height = 7, dpi = 300)

```



## Statistics


```{r}
# Extrae la tabla de abundancias (OTU table)
otu_table_data <- otu_table(all_data)

# Extrae la tabla de taxonomía
tax_table_data <- tax_table(all_data)
# Define los tres phylum de interés
phylum_of_interest <- c("Ascomycota", "Basidiomycota", "Microsporidia")

# Filtra la tabla de taxonomía por los tres phylum
subset_biom <- subset_taxa(all_data, Phylum %in% phylum_of_interest)
# Total de reads en todo el objeto
total_reads <- sum(otu_table(all_data))

# Total de reads en los tres phylum seleccionados
subset_reads <- sum(otu_table(subset_biom))

# Calcula el porcentaje de reads
porcentaje_reads <- (subset_reads / total_reads) * 100
porcentaje_reads

```



```{r}
#dataB

# Filtrar por un género específico
genero_of_interest <- "Acetilactobacillus"
subset_genero <- subset_taxa(dataB, Genus == genero_of_interest)

# Filtrar por el metadato `specie`
specie_of_interest <- "Melipona beecheii"
subset_specie <- subset_samples(dataB, Specie == specie_of_interest)

# Calcular el total de reads para esa especie
total_reads_specie <- sum(otu_table(subset_specie))

# Calcular el total de reads del género específico dentro de esa especie
subset_genero_specie <- subset_taxa(subset_specie, Genus == genero_of_interest)
genero_reads_specie <- sum(otu_table(subset_genero_specie))

# Calcular el porcentaje
porcentaje_genero <- (genero_reads_specie / total_reads_specie) * 100
porcentaje_genero
```


```{r}
#dataB
#10.56
# Filtrar por un género específico
genero_of_interest <- "Apilactobacillus"
subset_genero <- subset_taxa(dataB, Genus == genero_of_interest)

# Filtrar por el metadato `specie`
specie_of_interest <- "Scaptotrigona mexicana"
subset_specie <- subset_samples(dataB, Specie == specie_of_interest)

# Calcular el total de reads para esa especie
total_reads_specie <- sum(otu_table(subset_specie))

# Calcular el total de reads del género específico dentro de esa especie
subset_genero_specie <- subset_taxa(subset_specie, Genus == genero_of_interest)
genero_reads_specie <- sum(otu_table(subset_genero_specie))

# Calcular el porcentaje
porcentaje_genero <- (genero_reads_specie / total_reads_specie) * 100
porcentaje_genero
```