---
title: 'Test of Diversity of *Scaptotrigona Mexicana* and *Melipona beecheii*'
author:
 - Haydé Peruyero (haydeeperuyero@gmail.com)
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r, echo=FALSE, warning=FALSE}
library("phyloseq")
library("ggplot2")
library("RColorBrewer")
library("patchwork")
library("readr")
library("tidyverse")
library("grid")
library("gridExtra")
library("paletteer")
library("forcats")
```

First, we need to read the biom files. In this file, we found the following **Kingdoms**.

```{r, echo=FALSE}
# read the biom file
raw_metagenomes <- import_biom("D:/Users/hayde/Documents/R_sites/honey_diversity/data/honey_PF.biom")

# short the names, quit the first 4 characteres
raw_metagenomes@tax_table@.Data <- substring(raw_metagenomes@tax_table@.Data, 4)

# rename the columns
colnames(raw_metagenomes@tax_table@.Data)<- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
```


```{r, include=FALSE, message=FALSE, warning=FALSE}
# upload the metadata
df_meta <- read_csv("D:/Users/hayde/Documents/R_sites/honey_diversity/data/metadata_miel.csv")
```

```{r, include=FALSE}
# rename the names of the samples
# change n_specie to specie_n
df_meta <- df_meta %>%
  mutate(SampleID = case_when(
    grepl("_Scapto", ID) ~ paste0("Scapto_", sub("_Scapto.*", "", ID)),
    grepl("_Melli", ID) ~ paste0("Meli_", sub("_Melli.*", "", ID)),
    TRUE ~ as.character(ID) 
  ))
```


```{r, echo=FALSE}
# change the metadata name of samples to construct a new biom file
df_metadata <- sample_data(df_meta)
rownames(df_metadata) <- df_meta$ID
```


```{r, echo=FALSE}
# construc the new biom file with the new names of the samples
all_data <- merge_phyloseq(raw_metagenomes, sample_data(df_metadata))
```

```{r, echo=FALSE}
# Obtener los nombres de las muestras
sample_names <- sample_names(all_data)

# Cambiar nombres con "_Melli" a "M_número"
sample_names <- gsub("(\\d+)_Melli", "Meli_\\1", sample_names)

# Cambiar nombres con "_Scapto" a "S_número"
sample_names <- gsub("(\\d+)_Scapto", "Scapto_\\1", sample_names)

# Asignar los nuevos nombres de las muestras al objeto phyloseq
sample_names(all_data) <- sample_names

# Verificar los nuevos nombres de las muestras
sample_names(all_data)
```

```{r, echo = FALSE}
# the porcentajes of each kingdom
tax_table <- as.data.frame(tax_table(all_data))
reino_col <- "Kingdom"  # Ajusta el nombre de la columna si es diferente
reino_counts <- table(tax_table[[reino_col]])
total_observaciones <- sum(reino_counts)
porcentajes <- (reino_counts / total_observaciones) * 100
porcentajes
```


Obtain the quantity of reads in each sample.

```{r, echo=FALSE}
dprof <- data.frame(Samples = colnames(all_data@otu_table@.Data),
                    Reads = sample_sums(all_data),
                    Specie = all_data@sam_data@.Data[[15]])
#dprof
```



## Deep reads

In the following plot, we can see the deep of the reads for each sample in the honey of *Scaptotrigona mexicana* and *Melipona beecheii* specie.

```{r, echo=FALSE}
reads <- ggplot(data = dprof, mapping = aes(x = Samples, y = Reads))+
  geom_bar(stat = "identity", aes( fill = Specie)) +
  scale_fill_manual(values=c("#b8860b", "#FFBF00"))+
  labs(title = "Deep")+
  theme(axis.text.x = element_text(angle = 55, hjust = 1))+
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "bold"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14, face = "italic"), # Tamaño del texto de la leyenda
    legend.position = "bottom"
  )
  
reads
```


```{r, eval=FALSE, include=FALSE}
#ggsave(filename = "~/miel/Metagenomics/figures/deep_Scapto.svg", plot = reads, width = 8, height = 5, dpi = 300)

ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/deep_reads.svg", plot = reads, width = 10, height = 8, dpi = 300)
```

We will work with the `Bacteria` subset.

# Kigdoms

```{r, echo=FALSE}
# Resumen inicial de los datos faltantes
#summary(all_data@tax_table@.Data == "")

# Filtrar los datos eliminando Archaea y Homo sapiens
raw_metagenomes_all <- subset_taxa(
  all_data,
  Kingdom != "Archaea" & !(Kingdom == "Eukaryota" & Genus == "Homo")
)

# Resumen de datos después del filtrado
#summary(raw_metagenomes_all@tax_table@.Data == "")

# Agrupar por Kingdom (sin transformar a porcentajes)
abs_phylum_global <- tax_glom(raw_metagenomes_all, taxrank = 'Kingdom')

# Convertir a data.frame para visualización y análisis
abs_phyl_global_df <- psmelt(abs_phylum_global)

# Inspeccionar la estructura del data frame resultante
#str(abs_phyl_global_df)

# Convertir la columna 'Kingdom' en factor
abs_phyl_global_df$Kingdom <- as.factor(abs_phyl_global_df$Kingdom)
```
```{r, echo=FALSE}
# Define your custom color palette
phylum_colors_rel <- c(
  "Bacteria" = "#4F81BD", ##1f78b4", 
  "Eukaryota" = "#9BBB59"  , #"#33a02c",
  "Viruses" = "#F79646"#"#6a3d9a"
)

abs_phyl_global_df$Kingdom <- factor(abs_phyl_global_df$Kingdom, levels = c("Bacteria", "Viruses", "Eukaryota"))

relative_plot_phylum <- ggplot(data=abs_phyl_global_df, aes(x=Sample, y=Abundance, fill=Kingdom))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = phylum_colors_rel)+
  labs(x= "Samples", y = "Absolute abundance") +
  theme( legend.text = element_text(size = 16, face = "italic"), 
         legend.key.size = unit(0.4, 'cm'), 
         axis.title = element_text(size = 16), 
         axis.text.x = element_text(size = 16, angle = 90),
         axis.text.y = element_text(size = 16), 
         plot.title = element_text(size = 20), 
         axis.title.x = element_text(size = 16, face = "bold"),
         axis.title.y = element_text(size = 16, face = "bold"), 
         legend.title = element_text(size = 16),
         legend.position = "bottom")


relative_plot_phylum
```

```{r, eval=FALSE, include=FALSE}
ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/absolute_abundance_kingdom.svg", plot = relative_plot_phylum, width = 7, height = 5, dpi = 300)
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/relative_abundance_phylum.svg", plot = relative_plot_phylum, width = 6, height = 5, dpi = 300)
```

```{r, echo=FALSE}
# Resumen inicial de los datos faltantes
#summary(all_data@tax_table@.Data == "")

# Subconjunto eliminando Archaea y Homo sapiens
# Ajusta los nombres de las columnas de acuerdo con tu tax_table
raw_metagenomes_all <- subset_taxa(
  all_data,
  Kingdom != "Archaea" & !(Kingdom == "Eukaryota" & Genus == "Homo")
)

# Resumen de datos después del filtrado
#summary(raw_metagenomes_all@tax_table@.Data == "")

# Agrupar por Kingdom
rel_phylum_global <- tax_glom(raw_metagenomes_all, taxrank = 'Kingdom')

# Transformar los datos a porcentajes por muestra
rel_phylum_global <- transform_sample_counts(rel_phylum_global, function(x) x * 100 / sum(x))

# Convertir a data.frame para visualización y análisis
rel_phyl_global_df <- psmelt(rel_phylum_global)

# Inspeccionar la estructura del data frame resultante
#str(rel_phyl_global_df)

# Convertir la columna 'Kingdom' en factor
rel_phyl_global_df$Kingdom <- as.factor(rel_phyl_global_df$Kingdom)

```


```{r, echo=FALSE}
# Define your custom color palette
phylum_colors_rel <- c(
  "Bacteria" = "#4F81BD", ##1f78b4", 
  "Eukaryota" = "#9BBB59"  , #"#33a02c",
  "Viruses" = "#F79646"#"#6a3d9a"
)

rel_phyl_global_df$Kingdom <- factor(rel_phyl_global_df$Kingdom, levels = c("Bacteria", "Viruses", "Eukaryota"))

relative_plot_phylum <- ggplot(data=rel_phyl_global_df, aes(x=Sample, y=Abundance, fill=Kingdom))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = phylum_colors_rel)+
  labs(x= "Samples", y = "Relative abundance") +
  theme( legend.text = element_text(size = 16, face = "italic"), 
         legend.key.size = unit(0.4, 'cm'), 
         axis.title = element_text(size = 16), 
         axis.text.x = element_text(size = 16, angle = 90),
         axis.text.y = element_text(size = 16), 
         plot.title = element_text(size = 20), 
         axis.title.x = element_text(size = 16, face = "bold"),
         axis.title.y = element_text(size = 16, face = "bold"), 
         legend.title = element_text(size = 16),
         legend.position = "bottom")


relative_plot_phylum
```
```{r, eval=FALSE, include=FALSE}
ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/relative_abundance_kingdom.svg", plot = relative_plot_phylum, width = 7, height = 5, dpi = 300)
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/relative_abundance_phylum.svg", plot = relative_plot_phylum, width = 6, height = 5, dpi = 300)
```

# Bacteria

A summary of the Bacteria subset:


```{r, echo=FALSE}
# take a subset of the bacterias
data <- subset_taxa(all_data, Kingdom == "Bacteria")
#sample_sums(data)
#summary(data@otu_table@.Data)
```


## Alpha diversity 

- **Observed:** The Observed index simply counts the number of unique species (or Operational Taxonomic Units, OTUs) in a sample. It is a basic measure of species richness.

**Formula:**  
\[ S_{\text{obs}} = \sum_{i=1}^S I(n_i > 0) \]  
where \( S \) is the total number of species and \( I(n_i > 0) \) is an indicator function that is 1 if species \( i \) is observed (i.e., \( n_i > 0 \)) and 0 otherwise.

- **Chao1:** The Chao1 index is an estimator of the true species richness in a sample, accounting for the number of rare species. It is particularly useful when dealing with undersampled communities.

**Formula:**  
\[ \text{Chao1} = S_{\text{obs}} + \frac{F_1(F_1 - 1)}{2(F_2 + 1)} \]  
where \( F_1 \) is the number of singletons (species observed exactly once) and \( F_2 \) is the number of doubletons (species observed exactly twice).

- **Shannon:** The Shannon index, also known as Shannon-Wiener or Shannon-Weaver index, measures the diversity in a community by considering both the richness and the evenness of species. It accounts for the abundance of each species.

**Formula:**  
\[ H' = -\sum_{i=1}^S p_i \ln(p_i) \]  
where \( S \) is the total number of species and \( p_i \) is the proportion of individuals belonging to species \( i \).

In the case of the samples of honey, we can observed the following with respect to the diversity of these indexes. 

The Observed index counts the number of unique species in each sample. The plot shows that:
- Sample `23S` has the lowest observed species richness.
- Sample `92S` has the highest observed species richness.
- There is variability in species richness among the samples, with some samples having significantly higher species richness compared to others.

The Chao1 index estimates the true species richness, accounting for rare species. The plot indicates that:
- Sample `23S` has the lowest estimated species richness.
- Sample `92S` has the highest estimated species richness, similar to the Observed index.
- The error bars represent the uncertainty in the Chao1 estimates, showing that some samples have a higher level of uncertainty.
- The pattern of richness among samples is consistent with the Observed index, but Chao1 generally provides a higher estimate due to its correction for unseen species.

The Shannon index measures both species richness and evenness. The plot shows that:
- Sample `93S` has the lowest Shannon diversity, indicating either lower richness, lower evenness, or both.
- Sample `23S` has the highest Shannon diversity, suggesting higher richness, evenness, or both.
- The variability in Shannon diversity across samples is smaller compared to Observed and Chao1 indices, suggesting that evenness plays a significant role in these diversity measures.

In general, we can see the following:

- The samples exhibit different levels of species richness and diversity.
- Samples `23S` and `92S` stand out, with `23S` having the lowest richness but highest evenness, and `92S` having the highest richness but moderate evenness.
- Chao1 provides a more conservative estimate of species richness by accounting for rare species, which is reflected in the wider range and error bars.
- The Shannon index suggests that species evenness varies across samples, impacting the overall diversity.


```{r, echo=FALSE}
alpha_plot <- plot_richness(physeq = data, color= "Locality",
                                measures = c("Observed", "Shannon"))+
  geom_point(shape = 15, size = 3) +
  scale_color_manual(values = c("#009292", "#FF6DB6", "#006DDB", "#B66DFF", "#924900", "#DB6D00"))+
  labs(x= "Samples")+
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "bold"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14), # Tamaño del texto de la leyenda
    legend.position = "bottom"
  )
alpha_plot
```

```{r, eval=FALSE, include=FALSE}
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/alpha_div_plot_shanon_observed.svg", plot = alpha_plot, width = 8, height = 5, dpi = 300)

ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/alpha_div_plot_shanon_observed.svg", plot = alpha_plot, width = 7, height = 5, dpi = 300)
```

```{r, echo=FALSE}
alpha_plot <- plot_richness(physeq = data, color= "Locality",
                                measures = c("Chao1"))+
  scale_color_manual(values = c("#009292", "#FF6DB6", "#006DDB", "#B66DFF", "#924900", "#DB6D00"))+
  labs(x= "Samples")+
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "bold"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14) # Tamaño del texto de la leyenda
  )
alpha_plot
```


```{r, eval=FALSE, include=FALSE}
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/alpha_div_plot_chao.svg", plot = alpha_plot, width = 8, height = 5, dpi = 300)

ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/alpha_div_plot_chao.svg", plot = alpha_plot, width = 7, height = 5, dpi = 300)
```



```{r, include=FALSE, echo=FALSE}
#summary(data@tax_table@.Data== "")
data <- subset_taxa(data, Genus != "")
#summary(data@tax_table@.Data== "")
```


## Beta diversity


Beta diversity measures the differences in species composition between samples. The following plot is an NMDS (Non-metric Multidimensional Scaling) plot, which is commonly used to visualize beta diversity. 

### Interpretation of the NMDS Plot

1. **Axes Interpretation**:
   - **NMDS1** and **NMDS2** represent the two dimensions that best capture the differences in species composition among the samples.
   - The scales on these axes are arbitrary and do not have direct biological meaning. Instead, the relative positions of the points are what matter.

2. **Points Representation**:
   - Each point on the plot represents a sample.
   - The closer two points are to each other, the more similar their species compositions are.
   - Conversely, points that are farther apart indicate samples with more dissimilar species compositions.

3. **Clustering**:
   - If points form clusters, it indicates groups of samples that have similar species compositions.
   - In this plot, there seems to be some clustering, which suggests that certain samples have more similar communities.

4. **Stress Value (not shown)**:
   - Normally, NMDS plots include a stress value, which indicates how well the two-dimensional plot represents the multi-dimensional data. Lower stress values (typically <0.2) suggest a good representation.
   - Since the stress value is not shown here, it's important to be cautious in interpreting the distances, but generally, NMDS plots provide a useful visual summary.

### Overall Interpretation

- The NMDS plot shows the beta diversity among the samples, highlighting similarities and differences in their species compositions.
- Samples that are close to each other on the plot have similar species compositions, while those that are far apart are more dissimilar.
- Identifying clusters or gradients in the plot can help infer patterns related to environmental gradients, treatments, or other factors influencing the community composition.


In the plot we can see that there are some correlation between the samples and the city of the samples.

```{r, include=FALSE}
#Bacteria relative abundance
rel_data <- transform_sample_counts(data, function(x) x*100 / sum(x) )
```

```{r, echo=FALSE}
meta_ord <- ordinate(physeq = rel_data, method = "NMDS", 
                     distance = "bray")

beta_div_plot <- plot_ordination(physeq = rel_data, ordination = meta_ord, color="Locality", shape="Specie") +
  geom_point(size = 4) +  # Adjust point size
  scale_color_manual(values = c("#009292", "#FF6DB6", "#006DDB", "#B66DFF", "#924900", "#DB6D00"))+
  theme_minimal() +  # Use a minimal theme for clean look
  theme(legend.position = "right") +  # Position legend on the right
  labs(x = "NMDS1", y = "NMDS2")+
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "bold"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14, face = "italic") # Tamaño del texto de la leyenda
  )+
  guides(
    color = guide_legend(override.aes = list(shape = 15))
  )
beta_div_plot
```



```{r, eval=FALSE, include=FALSE}
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/beta_div_plot.svg", plot = beta_div_plot, width = 8, height = 8, dpi = 300)

ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/beta_div_plot.svg", plot = beta_div_plot, width = 8, height = 6, dpi = 300)
```


## Relative abundance

The relative abundance of taxa in metagenomic samples provides insights into the proportional representation of microbial groups within a given community. Relative abundance is calculated as the proportion of a specific taxon's abundance to the total abundance across all taxa in a sample. This metric is commonly expressed as a percentage or fraction.

For this study, we computed the relative abundance of bacterial taxa at the phylum and genus levels to identify dominant and rare microbial groups in the samples. High relative abundance values indicate predominant taxa that likely play significant roles in the community's ecological functions, while low relative abundance values suggest the presence of less dominant or transient taxa.

Visualization of relative abundance data was performed using bar plots, allowing for an intuitive comparison of microbial composition between samples and experimental conditions. These comparisons provide critical information about the dynamics of microbial populations and their potential functional implications.





### Predominant Phyla in Honey Samples

The analysis of microbial composition revealed that the phyla *Bacillota* and *Pseudomonadota* were predominant in the honey samples from both *M. beecheii* and *S. mexicana*. These phyla accounted for the majority of the microbial community across all samples, suggesting their critical role in shaping the microbiota of these honeys.

The dominance of Bacillota may be associated with its known members, including lactic acid bacteria, which contribute to the antimicrobial properties and preservation of honey. Similarly, Pseudomonadota includes taxa that could influence the physicochemical properties of honey through their metabolic activities. The consistent presence of these phyla in both honey types highlights their potential importance in the functional and ecological roles of stingless bee honeys.

Further analysis at the genus level is required to better understand the specific taxa contributing to these patterns and their potential functional implications.

```{r, include=FALSE}
#Subset
rel_phylum_global <- tax_glom(data, taxrank = 'Phylum')
rel_phylum_global <- transform_sample_counts(rel_phylum_global, function(x) x*100 / sum(x) )
rel_phyl_global_df<-psmelt(rel_phylum_global) # convert to data frame
#str(rel_phyl_global_df)
rel_phyl_global_df$Phylum[rel_phyl_global_df$Abundance < 5] <- "Phyla < 5% abund"
rel_phyl_global_df$Phylum <- as.factor(rel_phyl_global_df$Phylum)
unique(rel_phyl_global_df$Phylum)
```

```{r, echo=FALSE}
# Define your custom color palette
phylum_colors_rel <- c(
  "Bacillota" = "#0078A0", ##1f78b4", 
  "Pseudomonadota" = "#499878"  , #"#33a02c",
  "Phyla < 5% abund" = "#778899"#"#6a3d9a"
)

rel_phyl_global_df$Phylum <- factor(rel_phyl_global_df$Phylum, levels = c("Bacillota", "Pseudomonadota", "Phyla < 5% abund"))

relative_plot_phylum <- ggplot(data=rel_phyl_global_df, aes(x=Sample, y=Abundance, fill=Phylum))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = phylum_colors_rel)+
  labs(x= "Samples", y = "Relative abundance") +
  theme( legend.text = element_text(size = 16, face = "italic"), 
         legend.key.size = unit(0.4, 'cm'), 
         axis.title = element_text(size = 16), 
         axis.text.x = element_text(size = 16, angle = 90),
         axis.text.y = element_text(size = 16), 
         plot.title = element_text(size = 20), 
         axis.title.x = element_text(size = 16, face = "bold"),
         axis.title.y = element_text(size = 16, face = "bold"), 
         legend.title = element_text(size = 16),
         legend.position = "bottom")


relative_plot_phylum
```

```{r, eval=FALSE, include=FALSE}
ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/relative_abundance_phylum.svg", plot = relative_plot_phylum, width = 7, height = 5, dpi = 300)
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/relative_abundance_phylum.svg", plot = relative_plot_phylum, width = 6, height = 5, dpi = 300)
```

### Predominant Genera in Honey Samples

The microbial composition at the genus level showed clear distinctions between the honey samples of *M. beecheii* and *S. mexicana*. The predominant genera identified were *Acetilactobacillus*, *Apilactobacillus*, *Lactobacillus*, *Nicoliella*, *Pantoea*, *Tatumella*, and *Fructobacillus*.

*Acetilactobacillus* was the most abundant genus in the honey samples from *M. beecheii*, while *Apilactobacillus* dominated in the samples from *S. mexicana*. Interestingly, the genera *Tatumella* and *Pantoea* were only detected in two samples from *S. mexicana*, suggesting a more specific or transient presence in these honeys.

These results suggest distinct microbial signatures for each bee species, which may reflect differences in their foraging behaviors, hive environments, or nectar sources. The predominance of lactic acid bacteria such as *Acetilactobacillus*, *Apilactobacillus*, and *Lactobacillus* supports their potential role in the fermentation processes and preservation of honey.

Further investigation into the functional roles of these genera may provide insights into their contributions to the unique physicochemical properties of these stingless bee honeys.

```{r, include=FALSE}
#Subset
rel_Genus_global <- tax_glom(data, taxrank = 'Genus')
rel_Genus_global <- transform_sample_counts(rel_Genus_global, function(x) x*100 / sum(x) )
rel_Genus_global_df<-psmelt(rel_Genus_global) # convert to data frame
#str(rel_Genus_global_df)
rel_Genus_global_df$Genus[rel_Genus_global_df$Abundance < 5] <- "Genus < 5% abund"
rel_Genus_global_df$Genus <- as.factor(rel_Genus_global_df$Genus)
unique(rel_Genus_global_df$Genus)
```

```{r, echo=FALSE}
genus_colors <- c(
  "Acetilactobacillus" = "#002868", 
  "Apilactobacillus" = "#0077C2",
  "Lactobacillus" = "#66B2FF", 
  "Nicoliella" = "#ADD8E1",
  "Pantoea" = "#008080",
  "Tatumella" = "#98FB98",
  "Fructobacillus" = "#007BA7",
  "Genus < 5% abund" = "#778899"
)

rel_Genus_global_df$Genus <- factor(rel_Genus_global_df$Genus, levels = c("Acetilactobacillus", "Apilactobacillus", "Fructobacillus", "Lactobacillus", "Nicoliella", "Pantoea", "Tatumella","Genus < 5% abund" ))

relative_plot <- ggplot(data=rel_Genus_global_df, aes(x=Sample, y=Abundance, fill=Genus))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = genus_colors)+
  labs(x= "Samples", y = "Relative abundance") +
   theme(legend.text = element_text(size = 16, face = "italic"), 
         legend.key.size = unit(0.4, 'cm'), 
         axis.title = element_text(size = 16), 
         axis.text.x = element_text(size = 16, angle = 90),
         axis.text.y = element_text(size = 16), 
         plot.title = element_text(size = 20), 
         axis.title.x = element_text(size = 16, face = "bold"),
         axis.title.y = element_text(size = 16, face = "bold"), 
         legend.title = element_text(size = 16),
         legend.position = "bottom")

relative_plot
```

```{r, eval=FALSE, include=FALSE}
ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/relative_abundance_genus.svg", plot = relative_plot, width = 10, height = 8, dpi = 300)
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/relative_abundance_genus.svg", plot = relative_plot, width = 8, height = 7, dpi = 300)

```



# Virus

```{r, echo=FALSE}
# virus
data <- subset_taxa(all_data, Kingdom == "Viruses")
#sample_sums(data)
#summary(data@otu_table@.Data)

```


## Alfa diversity


```{r, echo=FALSE}
alpha_plot <- plot_richness(physeq = data, color= "Locality",
                                measures = c("Observed", "Shannon"))+
  geom_point(shape = 15, size = 3) +
  scale_color_manual(values = c("#009292", "#FF6DB6", "#006DDB", "#B66DFF", "#924900", "#DB6D00"))+
  labs(x= "Samples")+
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "bold"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14), # Tamaño del texto de la leyenda
    legend.position = "bottom"
  )
alpha_plot
```

```{r, eval=FALSE, include=FALSE}
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/alpha_div_plot_virus_shanon_observed.svg", plot = alpha_plot, width = 8, height = 5, dpi = 300)

ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/alpha_div_plot_virus_shanon_observed.svg", plot = alpha_plot, width = 7, height = 5, dpi = 300)
```

```{r, echo=FALSE}
alpha_plot_chao <- plot_richness(physeq = data, color= "Locality",
                                measures = c("Chao1"))+
  scale_color_manual(values = c("#009292", "#FF6DB6", "#006DDB", "#B66DFF", "#924900", "#DB6D00"))+
  labs(x= "Samples")+
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "bold"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14) # Tamaño del texto de la leyenda
  )
alpha_plot_chao
```


```{r, eval=FALSE, include=FALSE}
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/alpha_div_plot_virus_chao.svg", plot = alpha_plot_chao, width = 8, height = 5, dpi = 300)

ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/alpha_div_plot_virus_chao.svg", plot = alpha_plot_chao, width = 7, height = 5, dpi = 300)
```



## Beta diversity

```{r, include=FALSE, echo=FALSE}
#summary(data@tax_table@.Data== "")
data <- subset_taxa(data, Genus != "")
#summary(data@tax_table@.Data== "")
```

```{r, include=FALSE}
#Bacteria relative abundance
rel_data <- transform_sample_counts(data, function(x) x*100 / sum(x) )
```

```{r, echo=FALSE}
meta_ord <- ordinate(physeq = rel_data, method = "NMDS", 
                     distance = "bray")

beta_div_plot <- plot_ordination(physeq = rel_data, ordination = meta_ord, color="Locality", shape="Specie") +
  geom_point(size = 4) +  # Adjust point size
  scale_color_manual(values = c("#009292", "#FF6DB6", "#006DDB", "#B66DFF", "#924900", "#DB6D00"))+
  theme_minimal() +  # Use a minimal theme for clean look
  theme(legend.position = "right") +  # Position legend on the right
  labs(x = "NMDS1", y = "NMDS2")+
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "bold"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14, face = "italic") # Tamaño del texto de la leyenda
  )+
  guides(
    color = guide_legend(override.aes = list(shape = 15))
  )
beta_div_plot
```



```{r, eval=FALSE, include=FALSE}
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/beta_div_plot_virus.svg", plot = beta_div_plot, width = 8, height = 8, dpi = 300)

ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/beta_div_plot_virus.svg", plot = beta_div_plot, width = 8, height = 6, dpi = 300)
```


## Relative abundance


### Predominant Phyla in Honey Samples in Phylum


```{r, include=FALSE}
#Subset
rel_phylum_global <- tax_glom(data, taxrank = 'Phylum')
rel_phylum_global <- transform_sample_counts(rel_phylum_global, function(x) x*100 / sum(x) )
rel_phyl_global_df<-psmelt(rel_phylum_global) # convert to data frame
#str(rel_phyl_global_df)
rel_phyl_global_df$Phylum[rel_phyl_global_df$Abundance < 5] <- "Phyla < 5% abund"
rel_phyl_global_df$Phylum <- as.factor(rel_phyl_global_df$Phylum)
unique(rel_phyl_global_df$Phylum)
```

```{r, echo=FALSE}
# Define your custom color palette
phylum_colors_rel <- c(
  "Duplornaviricota" = "#1f78b4",  
  "Nucleocytoviricota" = "#a1d99b", 
  "Uroviricota" = "#fb6a4a",
  "Artverviricota" = "#c994c7",
  #"Kitrinoviricota" = "#fdae6b",
  "Phyla < 5% abund" = "#778899"
)

rel_phyl_global_df$Phylum <- factor(rel_phyl_global_df$Phylum, levels = c("Duplornaviricota", "Uroviricota", "Nucleocytoviricota", "Artverviricota", "Kitrinoviricota", "Phyla < 5% abund"))

relative_plot_phylum <- ggplot(data=rel_phyl_global_df, aes(x=Sample, y=Abundance, fill=Phylum))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = phylum_colors_rel)+
  labs(x= "Samples", y = "Relative abundance") +
  theme( legend.text = element_text(size = 16, face = "italic"), 
         legend.key.size = unit(0.4, 'cm'), 
         axis.title = element_text(size = 16), 
         axis.text.x = element_text(size = 16, angle = 90),
         axis.text.y = element_text(size = 16), 
         plot.title = element_text(size = 20), 
         axis.title.x = element_text(size = 16, face = "bold"),
         axis.title.y = element_text(size = 16, face = "bold"), 
         legend.title = element_text(size = 16),
         legend.position = "bottom")


relative_plot_phylum
```

```{r, eval=FALSE, include=FALSE}
ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/relative_abundance_phylum_virus.svg", plot = relative_plot_phylum, width = 10, height = 5, dpi = 300)
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/relative_abundance_phylum_virus.svg", plot = relative_plot_phylum_virus, width = 6, height = 5, dpi = 300)
```


### Predominant Genera in Honey Samples

```{r, include=FALSE}
#Subset
rel_Genus_global <- tax_glom(data, taxrank = 'Genus')
rel_Genus_global <- transform_sample_counts(rel_Genus_global, function(x) x*100 / sum(x) )
rel_Genus_global_df<-psmelt(rel_Genus_global) # convert to data frame
#str(rel_Genus_global_df)
rel_Genus_global_df$Genus[rel_Genus_global_df$Abundance < 5] <- "Genus < 5% abund"
rel_Genus_global_df$Genus <- as.factor(rel_Genus_global_df$Genus)
unique(rel_Genus_global_df$Genus)
```

```{r, echo=FALSE}
# Ordenar los géneros por abundancia total
rel_Genus_global_df <- rel_Genus_global_df %>%
  group_by(Genus) %>%
  mutate(TotalAbundance = sum(Abundance)) %>%
  ungroup()

# Ajustar los niveles del factor 'Genus' con "Genus < 5%" al final
rel_Genus_global_df$Genus <- fct_reorder(
  rel_Genus_global_df$Genus, 
  rel_Genus_global_df$TotalAbundance,
  .desc = TRUE
)

# Colocar "Genus < 5%" al final manualmente
rel_Genus_global_df$Genus <- fct_relevel(
  rel_Genus_global_df$Genus, 
  "Genus < 5% abund", 
  after = Inf
)

# Verificar los niveles ordenados
levels(rel_Genus_global_df$Genus)

```

```{r, echo=FALSE}
colors <- c('#778899','#006666', '#009999', '#00CCCC', '#00FFFF','#33FFFF','#65FFFF',
            '#99FFFF', '#B2FFFF', '#CBFFFF', '#E5FFFF', '#FFE5CB', '#FFCA99',
            '#FFAD65', '#FF8E33', '#FF6E00', '#CC5500','#D8152F','#A50021','#920000', '#993D00', '#662700')

relative_plot_g <- ggplot(data=rel_Genus_global_df, aes(x=Sample, y=Abundance, fill=Genus))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = colors)+
  labs(x= "Samples", y = "Relative abundance") +
   theme(legend.text = element_text(size = 16, face = "italic"), 
         legend.key.size = unit(0.4, 'cm'), 
         axis.title = element_text(size = 16), 
         axis.text.x = element_text(size = 16, angle = 90),
         axis.text.y = element_text(size = 16), 
         plot.title = element_text(size = 20), 
         axis.title.x = element_text(size = 16, face = "bold"),
         axis.title.y = element_text(size = 16, face = "bold"), 
         legend.title = element_text(size = 16),
         legend.position = "bottom")

relative_plot_g
```

```{r, eval=FALSE, include=FALSE}
ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/relative_abundance_genus_virus.svg", plot = relative_plot_g, width = 11, height = 8, dpi = 300)
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/relative_abundance_genus_virus.svg", plot = relative_plot_g, width = 8, height = 7, dpi = 300)

```



## Eukaryota


```{r, echo=FALSE}
# Eukaryota
data <- subset_taxa(all_data, Kingdom == "Eukaryota" & Genus != "Homo" )
#sample_sums(data)
#summary(data@otu_table@.Data)

```


```{r, echo=FALSE}
alpha_plot <- plot_richness(physeq = data, color= "Locality",
                                measures = c("Observed", "Shannon"))+
  geom_point(shape = 15, size = 3) +
  scale_color_manual(values = c("#009292", "#FF6DB6", "#006DDB", "#B66DFF", "#924900", "#DB6D00"))+
  labs(x= "Samples")+
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "bold"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14), # Tamaño del texto de la leyenda
    legend.position = "bottom"
  )
alpha_plot
```

```{r, eval=FALSE, include=FALSE}
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/alpha_div_plot_eukaryota_shanon_observed.svg", plot = alpha_plot, width = 8, height = 5, dpi = 300)

ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/alpha_div_plot_eukaryota_shanon_observed.svg", plot = alpha_plot, width = 7, height = 5, dpi = 300)
```

```{r, echo=FALSE}
alpha_plot_chao <- plot_richness(physeq = data, color= "Locality",
                                measures = c("Chao1"))+
  scale_color_manual(values = c("#009292", "#FF6DB6", "#006DDB", "#B66DFF", "#924900", "#DB6D00"))+
  labs(x= "Samples")+
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "bold"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14) # Tamaño del texto de la leyenda
  )
alpha_plot_chao
```


```{r, eval=FALSE, include=FALSE}
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/alpha_div_plot_eukaryota_chao.svg", plot = alpha_plot_chao, width = 8, height = 5, dpi = 300)

ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/alpha_div_plot_eukaryota_chao.svg", plot = alpha_plot_chao, width = 7, height = 5, dpi = 300)
```



## Beta diversity


```{r, include=FALSE}
#Bacteria relative abundance
rel_data <- transform_sample_counts(data, function(x) x*100 / sum(x) )
```

```{r, echo=FALSE}
meta_ord <- ordinate(physeq = rel_data, method = "NMDS", 
                     distance = "bray")

beta_div_plot <- plot_ordination(physeq = rel_data, ordination = meta_ord, color="Locality", shape="Specie") +
  geom_point(size = 4) +  # Adjust point size
  scale_color_manual(values = c("#009292", "#FF6DB6", "#006DDB", "#B66DFF", "#924900", "#DB6D00"))+
  theme_minimal() +  # Use a minimal theme for clean look
  theme(legend.position = "right") +  # Position legend on the right
  labs(x = "NMDS1", y = "NMDS2")+
  theme(
    text = element_text(size = 16), # Cambia el tipo de letra y tamaño
    plot.title = element_text(size = 16, face = "bold"), # Tamaño y estilo del título
    axis.title.x = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje X
    axis.title.y = element_text(size = 14, face = "bold"), # Tamaño y estilo del título del eje Y
    axis.text.x = element_text(size = 12), # Tamaño de las etiquetas del eje X
    axis.text.y = element_text(size = 12), # Tamaño de las etiquetas del eje Y
    legend.title = element_text(size = 14, face = "bold"), # Tamaño y estilo del título de la leyenda
    legend.text = element_text(size = 14, face = "italic") # Tamaño del texto de la leyenda
  )+
  guides(
    color = guide_legend(override.aes = list(shape = 15))
  )
beta_div_plot
```



```{r, eval=FALSE, include=FALSE}
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/beta_div_plot_eukaryota.svg", plot = beta_div_plot, width = 8, height = 8, dpi = 300)

ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/beta_div_plot_eukaryota.svg", plot = beta_div_plot, width = 8, height = 6, dpi = 300)
```

Existen 14 Phylums diferentes y dentro de esos solo 3 son Fungi. Se va a filtrar por solo esos.


```{r, echo=FALSE}
data <- subset_taxa(all_data, Kingdom == "Eukaryota" & Genus != "Homo" )
rel_phylum_global <- tax_glom(data, taxrank = 'Phylum')
rel_phylum_global <- transform_sample_counts(rel_phylum_global, function(x) x*100 / sum(x) )
rel_phyl_global_df<-psmelt(rel_phylum_global) # convert to data frame
#str(rel_phyl_global_df)
rel_phyl_global_df$Phylum <- as.factor(rel_phyl_global_df$Phylum)
unique(rel_phyl_global_df$Phylum)
```
```{r, echo=FALSE}
# Define your custom color palet

relative_plot_phylum <- ggplot(data=rel_phyl_global_df, aes(x=Sample, y=Abundance, fill=Phylum))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  #scale_fill_manual(values = phylum_colors_rel)+
  labs(x= "Samples", y = "Relative abundance") +
  theme( legend.text = element_text(size = 16, face = "italic"), 
         legend.key.size = unit(0.4, 'cm'), 
         axis.title = element_text(size = 16), 
         axis.text.x = element_text(size = 16, angle = 90),
         axis.text.y = element_text(size = 16), 
         plot.title = element_text(size = 20), 
         axis.title.x = element_text(size = 16, face = "bold"),
         axis.title.y = element_text(size = 16, face = "bold"), 
         legend.title = element_text(size = 16),
         legend.position = "bottom")


relative_plot_phylum
```


```{r, echo=FALSE}
data <- subset_taxa(data, Phylum == c("Mucoromycota", "Ascomycota", "Basidiomycota", "Microsporidia")) 
                       
#sample_sums(data)
#summary(data@otu_table@.Data)
```

```{r}
rel_phylum_global <- tax_glom(data, taxrank = 'Phylum')
rel_phylum_global <- transform_sample_counts(rel_phylum_global, function(x) x*100 / sum(x) )
rel_phyl_global_df<-psmelt(rel_phylum_global) # convert to data frame
#str(rel_phyl_global_df)
#rel_phyl_global_df$Phylum[rel_phyl_global_df$Abundance < 5] <- "Phyla < 5% abund"
rel_phyl_global_df$Phylum <- as.factor(rel_phyl_global_df$Phylum)
unique(rel_phyl_global_df$Phylum)
```


```{r, echo=FALSE}
# Define your custom color palette
phylum_colors_rel <- c(
  "Ascomycota" = "#1f78b4",  
  "Basidiomycota" = "#a1d99b",
  "Mucoromycota" = "#fb789b",
  "Microsporidia" = "#fb6a4a"
)

relative_plot_phylum <- ggplot(data=rel_phyl_global_df, aes(x=Sample, y=Abundance, fill=Phylum))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  scale_fill_manual(values = phylum_colors_rel)+
  labs(x= "Samples", y = "Relative abundance") +
  theme( legend.text = element_text(size = 16, face = "italic"), 
         legend.key.size = unit(0.4, 'cm'), 
         axis.title = element_text(size = 16), 
         axis.text.x = element_text(size = 16, angle = 90),
         axis.text.y = element_text(size = 16), 
         plot.title = element_text(size = 20), 
         axis.title.x = element_text(size = 16, face = "bold"),
         axis.title.y = element_text(size = 16, face = "bold"), 
         legend.title = element_text(size = 16),
         legend.position = "bottom")


relative_plot_phylum
```

```{r, eval=FALSE, include=FALSE}
ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/relative_abundance_phylum_eukaryota.svg", plot = relative_plot_phylum, width = 10, height = 5, dpi = 300)
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/relative_abundance_phylum_eukaryota.png", plot = relative_plot_phylum, width = 6, height = 5, dpi = 300)
```


### Predominant Genera in Honey Samples

```{r, include=FALSE}
#Subset
rel_Genus_global <- tax_glom(data, taxrank = 'Genus')
rel_Genus_global <- transform_sample_counts(rel_Genus_global, function(x) x*100 / sum(x) )
rel_Genus_global_df<-psmelt(rel_Genus_global) # convert to data frame
#str(rel_Genus_global_df)
rel_Genus_global_df$Genus[rel_Genus_global_df$Abundance < 5] <- "Genus < 5% abund"
rel_Genus_global_df$Genus <- as.factor(rel_Genus_global_df$Genus)
unique(rel_Genus_global_df$Genus)
```

```{r, echo=FALSE}
# Ordenar los géneros por abundancia total
rel_Genus_global_df <- rel_Genus_global_df %>%
  group_by(Genus) %>%
  mutate(TotalAbundance = sum(Abundance)) %>%
  ungroup()

# Ajustar los niveles del factor 'Genus' con "Genus < 5%" al final
rel_Genus_global_df$Genus <- fct_reorder(
  rel_Genus_global_df$Genus, 
  rel_Genus_global_df$TotalAbundance,
  .desc = TRUE
)

# Colocar "Genus < 5%" al final manualmente
rel_Genus_global_df$Genus <- fct_relevel(
  rel_Genus_global_df$Genus, 
  "Genus < 5% abund", 
  after = Inf
)

# Verificar los niveles ordenados
levels(rel_Genus_global_df$Genus)

```

```{r, echo=FALSE}

relative_plot_g <- ggplot(data=rel_Genus_global_df, aes(x=Sample, y=Abundance, fill=Genus))+ 
  geom_bar(aes(), stat="identity", position="stack")+
  #scale_fill_paletteer_d("colorBlindness::Blue2DarkOrange18Steps")+
  #scale_fill_manual(values = colors)+
  labs(x= "Samples", y = "Relative abundance") +
   theme(legend.text = element_text(size = 16, face = "italic"), 
         legend.key.size = unit(0.4, 'cm'), 
         axis.title = element_text(size = 16), 
         axis.text.x = element_text(size = 16, angle = 90),
         axis.text.y = element_text(size = 16), 
         plot.title = element_text(size = 20), 
         axis.title.x = element_text(size = 16, face = "bold"),
         axis.title.y = element_text(size = 16, face = "bold"), 
         legend.title = element_text(size = 16),
         legend.position = "bottom")

relative_plot_g
```

```{r, eval=FALSE, include=FALSE}
ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/relative_abundance_genus_eukaryota.svg", plot = relative_plot_g, width = 11, height = 8, dpi = 300)
#ggsave(filename = "D:/Users/hayde/Documents/R_sites/honey_diversity/figures/relative_abundance_genus_eukaryota.png", plot = relative_plot_g, width = 8, height = 7, dpi = 300)
```


# Statistics

Porcentaje de los Phylums: "Ascomycota", "Basidiomycota", "Microsporidia".

```{r, echo=FALSE}
# Extrae la tabla de abundancias (OTU table)
otu_table_data <- otu_table(all_data)

# Extrae la tabla de taxonomía
tax_table_data <- tax_table(all_data)
# Define los tres phylum de interés
phylum_of_interest <- c("Ascomycota", "Basidiomycota", "Microsporidia")

# Filtra la tabla de taxonomía por los tres phylum
subset_biom <- subset_taxa(all_data, Phylum %in% phylum_of_interest)
# Total de reads en todo el objeto
total_reads <- sum(otu_table(all_data))

# Total de reads en los tres phylum seleccionados
subset_reads <- sum(otu_table(subset_biom))

# Calcula el porcentaje de reads
porcentaje_reads <- (subset_reads / total_reads) * 100
porcentaje_reads

```


```{r, echo=FALSE}
# take a subset of the bacterias
data <- subset_taxa(all_data, Kingdom == "Bacteria")
#sample_sums(data)
#summary(data@otu_table@.Data)
```
Porcentaje del Género *Acetilactobacillus* en *Melipona beecheii*.

```{r}
#data Bacteria
#54.21

# Filtrar por un género específico
genero_of_interest <- "Acetilactobacillus"
subset_genero <- subset_taxa(data, Genus == genero_of_interest)

# Filtrar por el metadato `specie`
specie_of_interest <- "Melipona beecheii"
subset_specie <- subset_samples(data, Specie == specie_of_interest)

# Calcular el total de reads para esa especie
total_reads_specie <- sum(otu_table(subset_specie))

# Calcular el total de reads del género específico dentro de esa especie
subset_genero_specie <- subset_taxa(subset_specie, Genus == genero_of_interest)
genero_reads_specie <- sum(otu_table(subset_genero_specie))

# Calcular el porcentaje
porcentaje_genero <- (genero_reads_specie / total_reads_specie) * 100
porcentaje_genero
```
Porcentaje del género *Apilactobacillus* en *Scaptotrigona mexicana*.

```{r}
#dataB
#10.56
# Filtrar por un género específico
genero_of_interest <- "Apilactobacillus"
subset_genero <- subset_taxa(data, Genus == genero_of_interest)

# Filtrar por el metadato `specie`
specie_of_interest <- "Scaptotrigona mexicana"
subset_specie <- subset_samples(data, Specie == specie_of_interest)

# Calcular el total de reads para esa especie
total_reads_specie <- sum(otu_table(subset_specie))

# Calcular el total de reads del género específico dentro de esa especie
subset_genero_specie <- subset_taxa(subset_specie, Genus == genero_of_interest)
genero_reads_specie <- sum(otu_table(subset_genero_specie))

# Calcular el porcentaje
porcentaje_genero <- (genero_reads_specie / total_reads_specie) * 100
porcentaje_genero
```